---
title: "Homework 2"
author: "Joseph Froelicher"
date: October 6, 2021
output: pdf_document
header-includes:
  - \usepackage{amsmath}
---

```{r setup, echo = FALSE, include = FALSE, warning = FALSE}
library(janitor)
library(kableExtra)
library(lme4)
library(MASS)

# data
data = read.csv('../data/Cereal2.csv', header=TRUE) %>% clean_names()
data = data[data$fam_mem == 3,]

# for confidence intervals
alpha = 0.05
z = qnorm(1 - (alpha / 2))

```

\section{Question 1}
\subsection{Part A}

$$y_i\sim Pois(e^{\beta_{intercept} + \beta_{condition}x_{1i} + \beta_{sex}x_{2i}+\beta_{weight}x_{3i}})$$
$$Var(\textbf{Y}) = \mu$$

```{r question1_part_a, echo=TRUE, include=TRUE}
# part a (Poisson model)
model1 = glm(c1 ~ cond + factor(sex) + wt1, family = poisson(link = 'log'), data = data)
summary1 = summary(model1)
summary1$coefficients

```


Using a poisson regression model, in the presence of sex and weight, the odds of being in the experimental increase `r paste(round(exp(summary1$coefficients[2, 1]), 3), " (", round(exp(summary1$coefficients[2, 1] - summary1$coefficients[2, 2]*z), 3), ", ", round(exp(summary1$coefficients[2, 1] + summary1$coefficients[2, 2]*z), 3), ")", sep="")` times over those with no special treatment.

\subsection{Part B}
$$y_i\sim Pois(e^{\beta_{intercept} + \beta_{condition}x_{1i} + \beta_{sex}x_{2i}+\beta_{weight}x_{3i}})$$
$$Var(\textbf{Y}) = \phi\mu$$

```{r question1_part_b, echo=TRUE, include=TRUE}
# part b (Quasi-Poisson model)
model2 = glm(c1 ~ cond + factor(sex) + wt1, family=quasipoisson(link='log'), data=data)
summary2 = summary(model2)
summary2$coefficients

se1 = summary1$coefficients[2,2]
se1_by_dispersion = summary2$coefficients[2,2]/sqrt(summary2$dispersion)

```

Using a quasi-likelihood poisson regression model, in the presence of sex and weight, the odds of being in the experimental increase `r paste(round(exp(summary1$coefficients[2, 1]), 3), " (", round(exp(summary2$coefficients[2, 1] - summary2$coefficients[2, 2]*z), 3), ", ", round(exp(summary2$coefficients[2, 1] + summary2$coefficients[2, 2]*z), 3), ")", sep="")` times over those with no special treatment.

\subsection{Part C}

```{r question1_part_c, echo=TRUE, include=TRUE}
# part c (Random Normal Error Poisson model)
model3 = glmer(
  c1 ~ cond + factor(sex) + wt1 + (1 | fam_idno),
  family=poisson(link='log'),
  data=data
)
summary3 = summary(model3)
summary3$coefficients

```
$$y_i\sim Pois(e^{\beta_{intercept} + \beta_{condition}x_{1i} + \beta_{sex}x_{2i}+\beta_{weight}x_{3i}+\epsilon_i})$$
$$\epsilon_i\sim\mathcal{N}(0, \sigma^2)$$
Using a poisson regression model with random normal errors, in the presence of sex and weight, the odds of being in the experimental increase `r paste(round(exp(summary1$coefficients[2, 1]), 3), " (", round(exp(summary3$coefficients[2, 1] - summary3$coefficients[2, 2]*z), 3), ", ", round(exp(summary3$coefficients[2, 1] + summary3$coefficients[2, 2]*z), 3), ")", sep="")` times over those with no special treatment.


\subsection{Part D}
$$y_i\sim EF(\mu_i = log(\beta_{intercept} + \beta_{condition}x_{1i} + \beta_{sex}x_{2i}+\beta_{weight}x_{3i}), \phi)$$

```{r question1_part_d, echo=FALSE, include=TRUE}
# part d (Negative Binomial model)
model4 = glm.nb(c1 ~ cond + factor(sex) + wt1, link='log', data=data)
summary4 = summary(model4)
summary4$coefficients

```
Using a negative-binomial regression model, in the presence of sex and weight, the odds of being in the experimental increase `r paste(round(exp(summary4$coefficients[2, 1]), 3), " (", round(exp(summary4$coefficients[2, 1] - summary4$coefficients[2, 2]*z), 3), ", ", round(exp(summary4$coefficients[2, 1] + summary4$coefficients[2, 2]*z), 3), ")", sep="")` times over those with no special treatment.


\section{Question 2}
```{r table1, echo=FALSE, include=TRUE}
# part a
table1 = data.frame(
  " " = c(
    paste(round(exp(summary1$coefficients[1, 1]), 3)),
    paste(round(exp(summary1$coefficients[2, 1]), 3), " (", round(exp(summary1$coefficients[2, 1] - summary1$coefficients[2, 2]*z), 3), ", ", round(exp(summary1$coefficients[2, 1] + summary1$coefficients[2, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary1$coefficients[3, 1]), 3), " (", round(exp(summary1$coefficients[3, 1] - summary1$coefficients[3, 2]*z), 3), ", ", round(exp(summary1$coefficients[3, 1] + summary1$coefficients[3, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary1$coefficients[4, 1]), 3), " (", round(exp(summary1$coefficients[4, 1] - summary1$coefficients[4, 2]*z), 3), ", ", round(exp(summary1$coefficients[4, 1] + summary1$coefficients[4, 2]*z), 3), ")", sep=""),
    NA
  ),
  " " = c(
    paste(round(exp(summary2$coefficients[1, 1]), 3)),
    paste(round(exp(summary2$coefficients[2, 1]), 3), " (", round(exp(summary2$coefficients[2, 1] - summary2$coefficients[2, 2]*z), 3), ", ", round(exp(summary2$coefficients[2, 1] + summary2$coefficients[2, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary2$coefficients[3, 1]), 3), " (", round(exp(summary2$coefficients[3, 1] - summary2$coefficients[3, 2]*z), 3), ", ", round(exp(summary2$coefficients[3, 1] + summary2$coefficients[3, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary2$coefficients[4, 1]), 3), " (", round(exp(summary2$coefficients[4, 1] - summary2$coefficients[4, 2]*z), 3), ", ", round(exp(summary2$coefficients[4, 1] + summary2$coefficients[4, 2]*z), 3), ")", sep=""),
    round(summary2$dispersion, 3)
  ),
  " " = c(
    paste(round(exp(summary3$coefficients[1, 1]), 3)),
    paste(round(exp(summary3$coefficients[2, 1]), 3), " (", round(exp(summary3$coefficients[2, 1] - summary3$coefficients[2, 2]*z), 3), ", ", round(exp(summary3$coefficients[2, 1] + summary3$coefficients[2, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary3$coefficients[3, 1]), 3), " (", round(exp(summary3$coefficients[3, 1] - summary3$coefficients[3, 2]*z), 3), ", ", round(exp(summary3$coefficients[3, 1] + summary3$coefficients[3, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary3$coefficients[4, 1]), 3), " (", round(exp(summary3$coefficients[4, 1] - summary3$coefficients[4, 2]*z), 3), ", ", round(exp(summary3$coefficients[4, 1] + summary3$coefficients[4, 2]*z), 3), ")", sep=""),
    round(sqrt(summary3$varcor$fam_idno[1]), 3)
  ),
  " " = c(
    paste(round(exp(summary4$coefficients[1, 1]), 3)),
    paste(round(exp(summary4$coefficients[2, 1]), 3), " (", round(exp(summary4$coefficients[2, 1] - summary4$coefficients[2, 2]*z), 3), ", ", round(exp(summary4$coefficients[2, 1] + summary4$coefficients[2, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary4$coefficients[3, 1]), 3), " (", round(exp(summary4$coefficients[3, 1] - summary4$coefficients[3, 2]*z), 3), ", ", round(exp(summary4$coefficients[3, 1] + summary4$coefficients[3, 2]*z), 3), ")", sep=""),
    paste(round(exp(summary4$coefficients[4, 1]), 3), " (", round(exp(summary4$coefficients[4, 1] - summary4$coefficients[4, 2]*z), 3), ", ", round(exp(summary4$coefficients[4, 1] + summary4$coefficients[4, 2]*z), 3), ")", sep=""),
    round(summary4$theta, 3)
  )
)
rownames(table1) = c("Intercept", "Condition", "Sex", "Weight", "Other")
colnames(table1) = c("Poisson Regression", "Poisson QL", "Poisson + Normal error", "NB NLMIXED")
kable(table1, booktabs=TRUE)

```
\textbf{Table1}. Table of results from the four comparison models, Poisson, Quasi-Poisson, Random Normal Error Poisson, and Negative-Binomial.  
According to all our analyses (Poisson Regression, Quasi-Poisson Regression, Poisson Regression with Random Normal Error, and Negative Binomial Regression), the experimental group (cond=1) had much higher consumption. The Poisson and Quasi-Poisson Regression models both estimated that the experimental group was consuming `r round(exp(summary1$coefficients[2, 1]), 3)` times as much cereal. The normal random error model estimated that the experimental group was consuming `r round(exp(summary3$coefficients[2, 1]), 3)` times as much cereal, and the negative binomial model estimated that the experimental group was consuming `r round(exp(summary4$coefficients[2, 1]), 3)` times as much cereal, compared to the control group.
