---
title: "BIOS6643. L13 Generalized Linear Mixed Models"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
##library(tidyverse)
##library(ggplot2)
library(geepack)
library(lme4)


```



# GLMM for Seizure data 
## Epileptic Seizure Study of a randomized trial reported in Thall and Vail (1990). 
- 59 subjects with epilepsy suffering from simple or partial seizures were assigned at random to receive either the progabide drug  or a placebo 

- Number of seizures suffered by each subject over the 8-week period prior to the start of study was also recorded 

- After treatment initiation, the number of seizures for each subject was counted for each of 4 consecutive 2-week periods. 

```{r eval=T, echo = T, include=T}

#  Read in the data

dat.sz <- read.table("../data/epilepsy.txt")
colnames(dat.sz) <- c("subj","seize","visit","trt","base","age")
## trt=0 corresponds to placebo
head(dat.sz,3)

#  Create other covariates
dat.sz$o <- 8*(dat.sz$visit==0)+2*(dat.sz$visit>0)
dat.sz$logo <- log(dat.sz$o)
dat.sz$vm0 <- as.numeric(dat.sz$visit>0)


```

# Investigate if there is a different different effect after baseline visit 0
This means include an interaction between the indicator variable for visit>0 and the treatment indicator. 
```{r eval=T, echo = T, include=TRUE}
fit.glmm <- glmer(seize ~ offset(logo) + vm0*trt + (1 + vm0 | subj),
    family=poisson, data=dat.sz)

summary(fit.glmm) 

coef.glmm <-fixef(fit.glmm)
coef.glmm

```

The model we are fitting is 
$$\log (\mu_{ij})= \log(t_{ij}) + (\beta_1 + b_{b=1i}) + (\beta_2 vm0_{ij} + b_{2i} vm0_{ij}) + \beta_3 trt_i + \beta_4 trt_i * vm0_i, $$ 
where $t_{ij}=$ exposure time; $vm0=$ indicator for whether the visit is after baseline (1), $vm0=0$ for baseline visits; trt=1 if progabide and 0 if placebo.

Note interpretation of parameters is as follows

**Placebo** 

  - Baseline $\log (\mu_{ij}/T_{ij})= \beta_1 + b_{1i}$
  
  - Follow-up $\log (\mu_{ij}/T_{ij})= (\beta_1 + b_{1i}) + (\beta_{vm0} + b_{2i})$

**Progabide** 

  - Baseline $\log (\mu_{ij}/T_{ij})= \beta_1 + b_{1i} + \beta_{trt}$
  
  - Follow-up $\log (\mu_{ij}/T_{ij})= (\beta_1 + b_{1i}) + (\beta_{vm0} + b_{2i}) + \beta_{trt} + \beta_{vm0:trt}$

**Results**:

1. A patient treated with placebo has nearly the same expected seizure rate before and after randomization: exp ($\hat{\beta}_{vm0}=$) `r exp(coef.glmm[2])`

2. A patient treated with progabide has expected seizure rate reduced after treatment: exp($\hat{\beta}_{vm0} + \hat{\beta}_{vm0:trt}$) `r exp(coef.glmm[2] + coef.glmm[4])`

3. Estimated variance of the random intercepts and slopes is relatively large

# Marginal model
Interpret results of marginal model. 

```{r eval=T, echo = T, include=TRUE}
## AR1 
ar1.gee <- geeglm(seize ~ vm0*trt,id=subj,family=poisson("log"),
             offset=logo, corstr="ar1",data=dat.sz)
summary(ar1.gee)


```

# Investigate if there is a different effect after baseline visit 0 when adjusting for age in the model

```{r eval=T, echo = T, include=TRUE}
## AR1 
model1 = geeglm(
  seize ~ vm0*trt + age,
  id=subj,
  family=poisson("log"),
  offset=logo,
  corstr="ar1",
  data=dat.sz
)

summary(model1)


```
