---
title: "BIOS6643. L12 GEE and Marginal Models"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
##library(tidyverse)
##library(ggplot2)
library(gee)
library(geepack)


```



Fitting population-averaged models using the gee() function and  the geeglm() function in geepack.  We will need to install.packages("gee")
 and install.packages("geepack").

The syntax for each is similar.  A general call looks like

fit.object <- gee(model formula, id, corstr, family, data)

  - **corstr** is a specification for the working correlation matrix

  - **family** is a specification for the scaled exponential family that
 would be relevant under independence; the canonical link is the default

# Dental data
## Example of continuous outcome

```{r eval=T, echo = T, include=TRUE}

##  Read in the data 
dat.den <- read.csv("../data/dental.csv")
head(dat.den,3)


##dat.den$id <- as.factor(dat.den$id)
##dat.den$gender <- as.factor(dat.den$gender)

ind.gee.den <- gee(distance ~ age*gender,
             id = id,
             data = dat.den,
             family = gaussian,
             corstr = "independence"
             )
summary(ind.gee.den)
```

Recall the interpretation of the parameter estimates is at the population level. For instance for boys, the $\hat{\beta}_{age}=0.78$ may be interpreted as the average (expected) increase in distance per year of age increased.  


# Seizure data using gee()
## Epileptic Seizure Study of a randomized trial reported in Thall and Vail (1990). 
- 59 subjects with epilepsy suffering from simple or partial seizures were assigned at random to receive either the anti-epileptic drug progabide  or a placebo in addition to a standard chemotherapy regimen all were taking. 

- Because each individual might be prone to different rates of experiencing seizures, the investigators first tried to get a sense of this by recording the number of seizures suffered by each subject over the 8-week period prior to the start of administration of the assigned treatment. 

  - It is common in such studies to record such baseline measurements, so that the effect of treatment for each subject can be measured relative to how that subject behaved prior to treatment.

- Following initiation of treatment, the number of seizures for each subject was counted for each of 4 consecutive 2-week periods. 

- The age of each subject at the start of the study was also recorded, as it was suspected that subject age might be associated with the effect of the treatment.


```{r eval=T, echo = T, include=T}

#  Read in the data
dat.sz <- read.table("../data/epilepsy.txt")
colnames(dat.sz) <- c("subj","seize","visit","trt","base","age")
## trt=0 corresponds to placebo
head(dat.sz,3)

#  Create other covariates
dat.sz$o <- 8*(dat.sz$visit==0)+2*(dat.sz$visit>0)
dat.sz$logo <- log(dat.sz$o)
dat.sz$vm0 <- as.numeric(dat.sz$visit>0)

#  Basic models -- the unstructured fit using gee() does not converge
#  even with maxiter set to be much larger than the default

##un.gee <- gee(seize ~ trt + offset(logo), id=subj, family=poisson,
##              corstr="unstructured", data=dat.sz, maxiter=100)

cs.gee <- gee(seize ~ trt + offset(logo),id=subj,family=poisson,
              corstr="exchangeable", data=dat.sz)
summary(cs.gee)

## AR1 
ar1.gee <- gee(seize ~ trt + offset(logo),id=subj,family=poisson,
              corstr="AR-M",Mv=1,data=dat.sz)
 summary(ar1.gee)
 
```

# Seizure data using geeglm()

```{r eval=T, echo = T, include=TRUE}
## UN using geeglm
 un.geeglm <- geeglm(seize ~ trt, id=subj,family=poisson("log"),
             offset=logo, corstr="unstructured",data=dat.sz)
 summary(un.geeglm)

 ## CS using geeglm
 cs.geeglm <- geeglm(seize ~ trt,id=subj,family=poisson("log"),
             offset=logo, corstr="exchangeable",data=dat.sz)
summary(cs.geeglm)
 

ar1.geeglm <- geeglm(seize ~ trt,id=subj,family=poisson("log"),
             offset=logo, corstr="ar1",data=dat.sz)
summary(ar1.geeglm)

```

# Investigate if there is a different different effect after baseline visit 0
This means include an interaction between the indicator variable for visit>0 and the treatment indicator. Use two different working correlation structures and compare the results. 

```{r eval=T, echo=T, include=T}
baseline_interaction_model_ar1 = geeglm(
  seize ~ vm0*trt,
  id=subj,
  family=poisson('log'),
  offset=logo,
  corstr='ar1',
  data = dat.sz
)

summary(baseline_interaction_model_ar1)

baseline_interaction_model_exch = geeglm(
  seize ~ vm0*trt,
  id=subj,
  family=poisson('log'),
  offset=logo,
  corstr='exchangeable',
  data = dat.sz
)

summary(baseline_interaction_model_exch)

```
