---
title: "BIOS6643. L17  Joint Models of Longitudinal and Survival Data"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
##library(dplyr)
##library(tidyverse)
##library(ggplot2)

##library(HRW) ## to load spinal bone mineral data
library(nlme)
library(JM)


```
# In the JM package methods are available for the majority of the generic functions
- summary(), anova(), vcov(), logLik(), AIC() 
- coef(),fixef(),ranef()
- fitted(), residuals()
- plot()


## Primary Biliary Cirrhosis (PBC) study
- Primary biliary cirrhosis (PBC) is a chronic liver disease that leads to cirrhosis and eventually death

- 10-year study conducted by Mayo clinic (Murtagh et al., Hepatology, 1994) 

- 158 randomized to treatment, 154 to placebo

- Longitudinal biomarker measurements of serum bilirubin at times 0, 6m, 1y, 2y, etc.

\medskip
**Outcomes:**

  1. Longitudinal biomarker: serum bilirubin

  2. Time to death

**Question of interest:**

  - What is the association between the time-varying serum bilirubin (that is measured with error) and the risk of death?

```{r "pbc.data", eval=T, echo = T, include=T, out.width="90%"}
head(pbc2, 2)

```

## Joint model
```{r "jm", eval=T, echo = T, include=T, out.width="90%"}

lme.fit <- lme(log(serBilir) ~ year + year:drug, 
              data=pbc2, random = ~year|id)

surv.fit <- coxph(Surv(years, status2) ~ drug, 
              data=pbc2.id, x=TRUE)

## the knots of the piecewise constant are chosen based on the percentiles of events (5-6 internal knots by default)
joint.fit <- jointModel(lme.fit, surv.fit, 
              timeVar="year", method = "piecewise-PH-GH")
 
```

## Summary
```{r "jm.sum", eval=T, echo = T, include=T, out.width="90%"}
summary(joint.fit)

```

## Confidence intervals
```{r "jm.ci", eval=T, echo = T, include=T, out.width="90%"}
confint(joint.fit, parm='all')

```

## Comparison of nested models
Conduct a test for whether there is a drug effect in the survival model. 
```{r test, echo=TRUE, include=TRUE}
lme.fit0 <- lme(log(serBilir) ~ year + year:drug, 
              data=pbc2, random = ~year|id)


surv.fit0 <- coxph(Surv(years, status2) ~ 1, 
              data=pbc2.id, x=TRUE)

joint.fit0 <- jointModel(lme.fit0, surv.fit0, 
              timeVar="year", method = "piecewise-PH-GH")
 
anova(joint.fit0, joint.fit)

summary(joint.fit)

```

