---
title: "BIOS6643. L9 Specifying LMM through $G_i$ and $R_i$ structures"
cheauthor: "EJC"
-- date: "" 
header-includes:
   - \usepackage{amsmath}
   - \usepackage{float}
output: pdf_document
---

\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\itt}{\item}


```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- -->


```{r installp, echo = FALSE, eval=T, include=F}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(nlme)
library(lme4)

```

# Dental study

## Description

- The orthodontic study data of Potthoff and Roy (1964). 

- World famous data set that is used to introduce features of longitudinal data modeling and analysis

- A study was conducted involving 27 children, 16 boys and 11 girls

- For each child, the distance (mm) from the center of the pituitary to the pterygomaxillary fissure was measured at ages 8, 10, 12, and 14 years of age 

- The pterygomaxillary fissure is a vertical opening in the human skull.

- **Objectives** of the study included:

  - Determine if distances over time are larger on average for boys than for girls
  - Determine if the rate of change of distance over time is different for boys and girls.


```{r eval=T, echo = T, include=TRUE}
#  Read in the data 
dat.den <- read.table("../data/dental.txt")
dat.den <- dat.den[,2:5]      #  remove the first column
colnames(dat.den) <- c("id","age","distance","gender")

#  Total number of individuals
m <- max(dat.den$id)

head(dat.den, 2)
table(dat.den$gender)
```


## Some descriptives:  

```{r eval=T, echo = F, fig.height=3.7, include=T}
################################################################################################
### 

dat.den <- within(dat.den, {  id <- factor(id)
    gender <- factor(gender,levels=0:1,labels=c("Girls","Boys"))
})

ggplot(dat.den, aes(x = age, y = distance, group = id)) +
  geom_line(aes(color = gender), alpha = 0.5) +
  theme_classic() +
  theme(legend.position = "top") +
  ylab("Distance")

```



## a. Common G matrix for both genders with random intercept and slopes    

```{r eval=T, echo = T, include=T}
#  (a) Common G matrix for both genders, default diagonal within-child                               
#      covariance matrix R_i with same variance sigma^2 for each
#      gender.  

dental.lme.a <- lme(distance ~ -1 + gender + age:gender,data=dat.den,
                    random = ~ age | id,method="ML")
summary(dental.lme.a)
beta.a <- fixed.effects(dental.lme.a)  #  beta, also fixef(dental.lme.a)
b.a <- random.effects(dental.lme.a)    #  posterior modes bi, also ranef(dental.lme.a)
sebeta.a <- summary(dental.lme.a)$tTable[,"Std.Error"]  
## Recall we can get the var-cov of fixed coefficients using 'varFix'
## dental.lme.a$varFix

G.a <- getVarCov(dental.lme.a, type="random.effects")   #  G matrix
sigma2.a <- dental.lme.a$sigma^2  # sigma^2
V.a <- getVarCov(dental.lme.a, type="marginal", individual=1)   # V_i
R.a <- getVarCov(dental.lme.a, type="conditional", individual=1)   # R_i
G.a
V.a
R.a

```

## b. Common G matrix with random intercepts and slopes, diagonal within-child covariance matrix R_i with different variance for each gender         
```{r eval=T, echo = T, include=T}

dental.lme.b <- lme(distance ~ -1 + gender + age:gender,data=dat.den,
                    random = ~ age | id, weights = varIdent(form = ~ 1 | gender),
                    method="ML")
beta.b <- fixed.effects(dental.lme.b)  #  beta
sebeta.model.b <-  sqrt(diag(dental.lme.b$varFix))
b.b <- random.effects(dental.lme.b)    #  posterior modes bi
G.b <- getVarCov(dental.lme.b, type="random.effects")   #  G matrix
R.b.1 <- getVarCov(dental.lme.b,type="conditional",individual=1)   # R_1; first girl
R.b.12 <- getVarCov(dental.lme.b,type="conditional",individual=12)   # R_12; first boy
G.b
R.b.1
R.b.12

V.b.1 <- getVarCov(dental.lme.b,type="marginal",individual=1)   # V_1 (girl)
V.b.12 <- getVarCov(dental.lme.b,type="marginal",individual=12)   # V_12 (boy)
V.b.1
V.b.12

```


## c. Fit a common G matrix for both genders, and common within-child AR(1)
```{r eval=T, echo = T, include=T}
partc_model = lme(distance ~ -1 + gender + age:gender, data=dat.den, random = ~ age | id, correlation = corAR1(form = ~ age | id), method="ML")
summaryc = summary(partc_model)
R_c = getVarCov(partc_model1, type = "conditional", individual = 1)

```

## d. Fit a common G matrix for both genders, common within-child AR(1) exponential correlation
```{r eval=T, echo = T, include=T}
partd_model = lme(distance ~ -1 + gender + age:gender, data=dat.den, random = ~ age | id, correlation = corExp(form = ~ age | id), method="ML")
summaryd = summary(partd_model)
R_d = getVarCov(partd_model, type = "conditional", individual = 1)


```

## Compare models
```{r eval=T, echo = T, include=T}


```


## Refit final model using REML. Obtain population (PA) and subject-specific (SS) fitted values and residuals. 
**Note**: You may use *fitted(fit,level=0:1)* and *residuals(fit,level=0:1)*, where level=0 gives PA, level = 1 gives SS (default), and level = 0:1 gives both.  
```{r eval=T, echo = T, include=T}
partc_model = lme(distance ~ -1 + gender + age:gender, data=dat.den, random = ~ age | id, correlation = corAR1(form = ~ age | id), method="ML")
summaryc = summary(partc_model)
R_c = getVarCov(partc_model1, type = "conditional", individual = 1)

partd_model = lme(distance ~ -1 + gender + age:gender, data=dat.den, random = ~ age | id, correlation = corExp(form = ~ age | id), method="ML")
summaryd = summary(partd_model)
R_d = getVarCov(partd_model, type = "conditional", individual = 1)

```

# Fit a LMM with random intercept and slopes using lme4:lmer


